Supplement: Autopolyploid establishment through polygenic adaptation

### 2024-06-06 doc/paper/supp.tex
update

\subsection{Mixed-ploidy infinitesimal model}

\subsubsection{Discrete locus model and variance scaling \label{sec:scaling}}

Consider an $L$-locus additive model, with two alleles (0 and 1) at each locus.
For a $k$-ploid individual, let $X_{i,j}$ be the allele at homolog $j$ of locus
$i$.
We assume the trait value is determined by
\begin{equation}
  z = \sum_{i=1}^L\sum_{j=1}^k a_{i,k} X_{i,j}
\end{equation}
Where $a_{i,k}$ is the allelic effect of the 1 allele at locus $i$ in
$k$-ploids.
The genetic variance at HWLE in $k$-ploids ($\tilde{V}_{z,k}$) will then be
\begin{equation}
  \tilde{V}_{z,k} = k\sum_{i=1} a_{i,k}^2 p_iq_i = kV_{x,k}
\end{equation}
where we refer to $V_{x,k}$ as the variance associated with a haploid genome in
$k$-ploids at HWLE.
Note that we also have $\tilde{V}_{z,2} = 2V_{x,k} = 2V$, where $V$ is the
segregation variance in the diploid population.

We now assume $a_{i,k} = \beta_k a_{i,2}$, i.e. allelic effects in $k$-ploids
are as in diploids, but scaled by a factor $\beta_k$.
This implies that
\begin{equation}
  \frac{\tilde{V}_{z,k}}{\tilde{V}_{z,2}} 
  = \frac{kV_{x,k}}{2V_{x,2}} = \frac{k}{2} \beta_{k}^2
\end{equation}
and hence also that $V_{x,k} = \beta_k^2 V_{x,2} = \beta_k^2 V$.
These relations will also hold in the infinitesimal limit.
Below, we derive the segregation variance expressions for the different meiotic
processes in the mixed-ploidy models in terms of the $V_{x,k}$.
Using the relationships outlined here, we will then be able to express all the
required variances in terms of $V$.

Note that under the above assumptions, not only the variance is rescaled, but
also the expected value of offspring within a family.
Specifically, for a $k$-ploid offspring of parental pair $i$ and $j$, where the
gamete coming from $i$ is $g_i$-ploid and the other gamete $g_j$-ploid, we have
\begin{equation}
  \Ex[Z_{ij}] = \beta_k \left(\frac{g_i}{c_i}\frac{z_i}{\beta_{c_i}} + 
    \frac{g_j}{c_j}\frac{z_j}{\beta_{c_j}}\right)
\end{equation}
where $c_i$ is the ploidy level of individual $i$.
Here we rescale the parental trait values to the diploid scale by dividing by
$\beta_{c_i}$, sum the expected rescaled gametic values, and rescale to the
offspring ploidy level.

\subsubsection{Segregation variance \label{sec:segvar}}

@@

As an example, consider ordinary diploid meiosis at a single locus, where a
diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
gene from the population and $v_x$ is the variance in additive genetic values
at the locus.
We have
  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
Using \cref{eq:eve}, we can then find the segregation variance
  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
Summing over $L$ independent loci, we find the gametic segregation variance in
diploids (which is halve the zygotic segregation variance $V$, by definition) as
  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{V_{x,2}}{2}$$
and hence $V=V_{x,2}$, where $V_{x,2}$ is as defined in \cref{sec:scaling}.
This will hold in the infinitsimal limit, where $L$ becomes very large and
$v_x$ smaller and smaller.

Below, we derive the segregation variance associated with haploid and diploid
gamete production in the three cytotypes of the mixed-ploidy population.
We use \cref{eq:eve} to do this, and express the segregation variance for a
$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
variance $V_{x,k}$ associated with a haploid genome in a (hypothetical)
$k$-ploid reference population at HWLE. Using the scaling assumptions above,
i.e. $V_{x,k} = \beta_k^2 V$, we can then express all segregation variances in
terms of the diploid segregation variance in the reference population $V$.

@@

When double reduction occurs, an $ABCD$ genotype would generate 10 distinct
gametes, as opposed to 6 in when chromosomes form bivalents. As a result, the
segregation variance is increased by double reduction.
For a random genotype $X_1X_2X_3X_4$, we can find the gametic segregation
variance contributed by a locus when double reduction happens as
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
        &= \Var[2X] - \Var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
        &= 4v_x - \frac{1}{4}4v_x 
        = 3v_x
\end{align*}
where $X$ denotes the additive effect of a random allele at the locus drawn from the reference
population and $v_x = \Var[X]$.
In the absence of double reduction we have
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] 
        &= 2\Var[X] - \Var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
        = v_x
\end{align*}
Assuming that the probability of double reduction at any locus is $\alpha_4$,
and summing over independent loci, we find that the gametic segregation
variance in the presence of double reduction should be
\begin{equation}
V_{S(4,2)} = (1-\alpha_4)V_{x,4} + 3\alpha_4 V_{x,4} = V_{x,4}(1+2\alpha_4).
\label{eq:dr}
\end{equation}
where, again, $V_{x,4}$ is as defined in \cref{sec:scaling}.

@@

The mechanisms of unreduced gamete formation do not necessarily lead to a
faithful transmission of the complete diploid genome.
Unreduced gametes are formed in two ways, depending on the meiotic abberation
that leads to their origin: (1) first division restitution (FDR) of (2) second
division restitution (SDR) \citep{bretagnolle1995,storme2013}.
Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
recombination happens with probability $c$ and that conditional on unreduced
gamete formation, formation is due to FDR with probability $f$ while it is due
to SDR with probability $1-f$.
The different unreduced gametes that are formed are represented schematically
in \cref{fig:udip}.
Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
locus will be
$$Y = \begin{cases}
    2X_1 & \text{w.pr.  } \alpha_2/2 \\
    2X_2 & \text{w.pr.  } \alpha_2/2 \\
    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
    \end{cases}$$
where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
of the same gene end up in a diploid gamete produced by a diploid individual
(see the diagram in \cref{fig:udip}). 
Conditional on the latter event, we get
\begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*}
Conditioning on the complementary event, all gametes have genetic value
$Y = X_1+X_2$, so that the segregation variance is 0.
Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 V_{x,2}$.

@@

When a triploid produces a haploid gamete, we get as segregation variance at a
single locus
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= v_x - \Var\left[\frac{1}{3}\left(X_1 + X_2 + X_3\right)\right] \\ 
    &= v_x - \frac{1}{9}3v_x = \frac{2}{3}v_x
\end{align*}
Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}V_{x,3}$.
When a triploid produces a diploid gamete, we assume there is, as in diploids
and tetraploids, a probability $\alpha_3$ that the same gene copy ends up twice
in the gamete.
Conditional on this happening, we have
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= \Var[2X] - \Var\left[\frac{1}{3}\left(2X_1 + 2X_2 + 2X_3\right)\right] \\
    &= 4V_x - \frac{4}{9}3V_x = \frac{8}{3}V_x
\end{align*}
Conditional on this \textit{not} happening, 
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= 2V_x - \Var\left[\frac{1}{3}\left((X_1+X_2) + (X_1+X_3) +
      (X_2+X_3)\right)\right] \\
    &= 2V_x - \frac{4}{9}3V_x = \frac{2}{3}V_x
\end{align*} 
Putting this together and summing over many loci, the segregation variance for
a diploid gamete from a triploid individual would be
  $$V_{S(3,2)} = \frac{8}{3}V_{x,3}\alpha_3 + \frac{2}{3}V_{x,3} (1-\alpha_3) 
            = \frac{2}{3}V_{x,3}(1 + 3\alpha_3)$$

>>>>

\subsection{Segregation variance expressions for the mixed-ploidy infinitesimal model \label{sec:segvar}}

@@

As an example, consider ordinary diploid meiosis at a single locus, where a
diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
gene from the population and $v_x$ is the variance in additive genetic values
at the locus.
We have
  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
Using \cref{eq:eve}, we can then find the segregation variance
  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
Summing over $L$ independent loci, we find the gametic segregation variance in
diploids (which is halve the zygotic segregation variance $V$, by definition) as
  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{\Vx_{2}}{2}$$
and hence $V=\Vx_{2}$, where $\Vx_{2}$ is the genetic variance associated with
a haploid genome in the diploid cytotype, as defined in the main text.
This will hold in the infinitsimal limit, where $L$ becomes very large and
$v_x$ smaller and smaller.

Below, we derive the segregation variance associated with haploid and diploid
gamete production in the three cytotypes of the mixed-ploidy population.
We use \cref{eq:eve} to do this, and express the segregation variance for a
$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
variance $\Vx_{k}$ associated with a haploid genome in a (hypothetical)
$k$-ploid reference population at HWLE. Using the scaling assumptions outlined
in the main text, i.e. $\Vx_{k} = \beta_k^2 V$, we can then express all
segregation variances in terms of the diploid segregation variance in the
reference population $V$.

@@

When double reduction occurs, an $ABCD$ genotype would generate 10 distinct
gametes, as opposed to 6 in when chromosomes form bivalents. As a result, the
segregation variance is increased by double reduction.
For a random genotype $X_1X_2X_3X_4$, we can find the gametic segregation
variance contributed by a locus when double reduction happens as
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3,X_4]] \\
        &= \Var[2X] - \Var\left[\frac1 4 (2X_1 + 2X_2 + 2X_3 + 2X_4)\right] \\
        &= 4v_x - \frac{1}{4}4v_x 
        = 3v_x
\end{align*}
where $X$ denotes the additive effect of a random allele at the locus drawn from the reference
population and $v_x = \Var[X]$.
In the absence of double reduction we have
\begin{align*}
    \Ex[\Var[Y|X_1,X_2,X_3,X_4]] 
        &= 2\Var[X] - \Var\left[\frac1 6 \sum_{i=1}^3 \sum_{j=i+1}^4(X_i+X_j)\right]
        = v_x
\end{align*}
Assuming that the probability of double reduction at any locus is $\alpha_4$,
and summing over independent loci, we find that the gametic segregation
variance in the presence of double reduction should be
\begin{equation}
V_{S(4,2)} = (1-\alpha_4)\Vx_{4} + 3\alpha_4 \Vx_{4} = \Vx_{4}(1+2\alpha_4).
\label{eq:dr}
\end{equation}
where, again, $\Vx_{4}$ is the genetic variance associated with a haploid
genome in the (hypothetical) tetraploid reference population, as defined in the
main text.

@@

The mechanisms of unreduced gamete formation do not necessarily lead to a
faithful transmission of the complete diploid genome.
Unreduced gametes are formed in two ways, depending on the meiotic abberation
that leads to their origin: (1) first division restitution (FDR) of (2) second
division restitution (SDR) \citep{bretagnolle1995,storme2013}.
Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
recombination happens with probability $c$ and that conditional on unreduced
gamete formation, formation is due to FDR with probability $f$ while it is due
to SDR with probability $1-f$.
The different unreduced gametes that are formed are represented schematically
in \cref{fig:udip}.
Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
locus will be
$$Y = \begin{cases}
    2X_1 & \text{w.pr.  } \alpha_2/2 \\
    2X_2 & \text{w.pr.  } \alpha_2/2 \\
    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
    \end{cases}$$
where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
of the same gene end up in a diploid gamete produced by a diploid individual
(see the diagram in \cref{fig:udip}). 
Conditional on the latter event, we get
\begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*}
Conditioning on the complementary event, all gametes have genetic value
$Y = X_1+X_2$, so that the segregation variance is 0.
Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 \Vx_{2}$.

@@

When a triploid produces a haploid gamete, we get as segregation variance at a
single locus
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= v_x - \Var\left[\frac{1}{3}\left(X_1 + X_2 + X_3\right)\right] \\ 
    &= v_x - \frac{1}{9}3v_x = \frac{2}{3}v_x
\end{align*}
Summing over many loci, we have $V_{S(3,1)} = \frac{2}{3}\Vx_{3}$.
When a triploid produces a diploid gamete, we assume there is, as in diploids
and tetraploids, a probability $\alpha_3$ that the same gene copy ends up twice
in the gamete.
Conditional on this happening, we have
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= \Var[2X] - \Var\left[\frac{1}{3}\left(2X_1 + 2X_2 + 2X_3\right)\right] \\
    &= 4V_x - \frac{4}{9}3V_x = \frac{8}{3}V_x
\end{align*}
Conditional on this \textit{not} happening, 
\begin{align*}
  \Ex[\Var[Y|X_1,X_2,X_3]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2,X_3]] \\
    &= 2V_x - \Var\left[\frac{1}{3}\left((X_1+X_2) + (X_1+X_3) +
      (X_2+X_3)\right)\right] \\
    &= 2V_x - \frac{4}{9}3V_x = \frac{2}{3}V_x
\end{align*} 
Putting this together and summing over many loci, the segregation variance for
a diploid gamete from a triploid individual would be
  $$V_{S(3,2)} = \frac{8}{3}\Vx_{3}\alpha_3 + \frac{2}{3}\Vx_{3} (1-\alpha_3) 
            = \frac{2}{3}\Vx_{3}(1 + 3\alpha_3)$$
